---
title: "r_tutorial_week_01"
subtitle: "Werkcollege 1"
author: "Sebastian A. J. Lortz"
date: today
date-format: "D MMMM YYYY"
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    theme: flatly
---

```{r setup, include=FALSE}
# This block sets defaults for the whole document. Students can ignore it.
knitr::opts_chunk$set(
  echo = TRUE,        # show the code
  collapse = TRUE,    # keep code + output together
  comment = "#>",     # output lines start with #>
  prompt = FALSE
)
```

```{=html}
<style>
.your-turn {
  background-color: #eaf4f1;
  border-left: 5px solid #18bc9c;
  padding: 0.6em 1em 0.6em 1em;
  margin: 1em 0 1em 0;
  border-radius: 4px;
}
.your-turn h3 { margin-top: 0; }
</style>
```

# How to use this tutorial {.unnumbered}

Welcome to the R tutorial! You do **not** need any prior experience. Work through this page carefully, from top to bottom. Every short explanation is followed by a **Your turn** task to do yourself.

General tips:

-   Read each part slowly and in order. Each step builds on the one before it.
-   Type the code yourself instead of copying and pasting it. You learn faster this way.
-   Errors are normal. Read the message and try again.
-   If you are stuck, ask a teacher.
-   Save your work often.

------------------------------------------------------------------------

# Your computer's filing system

Before we start with RStudio, we discuss some important terms about your computer itself.

A **file** is a single object (a photo, a Word document, an R script). A **folder** (also called a **directory**) is a box that holds files and other folders. Folders can sit inside other folders, building a **folder structure**.

::: {.callout-note title="Example: a tidy folder structure"}
A tidy folder structure can look like this:

```         
Documents
└── first_year
    └── research_methods
        ├── data
        ├── scripts
        └── output
```
:::

At the root of this structure is a **Drive**. Thus, a drive is the storage of your computer. On a personal computer you usually have only one drive. On top of the drive sit some common **base folders**, such as the **Desktop** (the screen you see when your computer starts), **Documents** (where files are usually stored), and **Downloads** (where files you download are saved).

It is important to keep this structure tidy. Work inside **Documents**, make one folder for your first-year studies. Inside that folder, make a folder for each course. Inside each course folder, you can sort your material further.

Do **not** build your folder structure inside the **Downloads** folder. Everything you download lands there automatically, so it quickly becomes messy.

::: your-turn
### Your turn {.unnumbered}

Create one folder for this whole course, so everything stays in one place.

1.  Open your file explorer (*Windows Explorer* on Windows, *Finder* on Mac).
2.  Go to your **Documents** folder.
3.  If it is not there already, create a folder for your studies and name it `first_year`.
4.  Inside `first_year`, create a **new folder** and name it `research_methods`.
5.  Fill the folder `research_methods` with the subfolders in the example above.
:::

**Good practice for names:**

-   Keep folder and file names simple: use lowercase letters, numbers, and underscores `_` instead of spaces. For example, `my data.csv` can cause problems later, but `my_data.csv` does not.
-   Use short but clear names, so you know what is inside without opening the folder.
-   When a folder starts to hold many different things, make subfolders to group them.

You now have a home for your course work.

------------------------------------------------------------------------

# Opening RStudio

**R** is the engine that does the calculations. **RStudio** is the user interface you use to talk to that engine comfortably. You installed both before class (see the Brightspace instructions). We always open **RStudio** only and **never** open **R** itself.

::: your-turn
### Your turn {.unnumbered}

Open the **RStudio** program (not plain "R").
:::

::: {.callout-important title="Stuck on installation?"}
Raise your hand right away. A teacher will help you.
:::

## The four panes

RStudio's window is split into panes.

-   **Editor** (top-left): where you *write and save* code (empty for now).
-   **Console** (bottom-left): where R *runs* commands and shows results.
-   **Environment** (top-right): a list of everything you have stored (empty for now).
-   **Files / Plots / Packages / Help** (bottom-right): your file browser, graphs, installed packages, and help pages.

Find the **Console**. You will see a `>` symbol. That `>` means "R is ready and waiting for you." If, while programming, this symbol is not shown, click in the console and press *Esc* (escape) on your keyboard to get the `>` symbol back.

------------------------------------------------------------------------

# Working Directory

RStudio needs to know where to work, for example, where to look for files and data. The first thing we do when working in RStudio is to set the **working directory**, that is the folder we are working in. Because we already made a folder for this course, we choose that one.

::: your-turn
### Your turn {.unnumbered}

1.  In the menu, click *Session* \> *Set Working Directory* \> *Choose Directory*.
2.  A window opens that shows your computer's file system. Navigate to the `research_methods` folder you made and select it.
:::

RStudio has now set the working directory to that folder.

------------------------------------------------------------------------

# R scripts

In this course we only work with **R scripts**, which are shown in the **Editor** pane. An R script is a place where you type and save lines of code, which are then run. A script runs from top to bottom.

We work with R scripts because your progress in RStudio is lost when you close the program. But if you save your R script, you can open it again next time and run each line to get back to where you stopped last time.

This is important for **reproducibility**: you can give your script to another person, and they get the same output. We never type our code straight into the console.

::: your-turn
### Your turn {.unnumbered}

Create a new R script: click *File* \> *New File* \> *R Script*.
:::

In a script, we write notes with a `#` (hash) at the start of a line. A line that starts with `#` is called a **comment**. RStudio does not run a comment as code; it skips it. Comments help you and others understand what your code does.

::: your-turn
### Your turn {.unnumbered}

Click in the first line of your new R script. Type a `#` and then a title for your script, for example: `# R tutorial - Week 1`.
:::

------------------------------------------------------------------------

# R as a calculator

The fastest way to meet R is to use it as a calculator.

::: your-turn
### Your turn {.unnumbered}

Type each line below into your R script. Press *Enter* after each one, so the lines sit underneath each other.
:::

```{r}
2 + 3
10 - 4
6 * 7
20 / 5
```

To run a line, click somewhere in that line and:

-   on Windows, hold *Ctrl* and press *Enter*;
-   on Mac, hold *Cmd* and press *Enter*.

R runs the line and jumps to the next one.

::: your-turn
### Your turn {.unnumbered}

Run the four calculations above, one by one.
:::

R prints the answer back. Ignore the `[1]` in front for now -- it just means "this is the first result."

R also knows powers and square roots:

```{r}
2 ^ 3    # 2 to the power of 3
sqrt(9)  # the square root of 9
```

Remember: anything after a `#` is a **comment**: a note for humans that R ignores. Use comments to remind yourself what code does.

::: your-turn
### Your turn {.unnumbered}

In your script, work out:

(a) `144` divided by `12`, and
(b) `5` to the power of `2`.

*(Answers: `12` and `25`.)*
:::

------------------------------------------------------------------------

# Objects: storing values for later

Re-typing numbers every time is tedious. Instead we store a value in an **object** (sometimes called a *variable*) and give it a name. We do this with the **assignment arrow** `<-`, which you read as "gets."

```{r}
age <- 19
```

Read that as: "the object `age` gets the value 19." Notice nothing was printed. The value is now stored. To see it, type the name on its own:

```{r}
age
```

::: {.callout-tip title="Shortcut for the assignment arrow"}
Press *Alt* and *-* (minus) on Windows, or *Option* and *-* on Mac. RStudio types `<-` with the spaces for you.
:::

Once stored, an object can be used in calculations just like the number itself:

```{r}
self_efficacy <- 32   # a score on the self-efficacy scale we will collect
self_efficacy + 8
```

You can also **overwrite** an object by assigning a new value to the same name:

```{r}
age <- 20   # age is now 20, the old value (19) is gone
age
```

Check the **Environment** pane (top-right): `age` and `self_efficacy` now appear there. That pane is your live list of stored objects.

**Naming rules** (important, R is strict):

-   Names may use letters, numbers, `.` and `_`, and must **start with a letter**.
-   No spaces. Use `self_efficacy`, not `self efficacy`.
-   Names are **case-sensitive**: `Age`, `age`, and `AGE` are three different objects.

```{r}
#| error: true
Age   # we never created "Age" with a capital A, so R complains
```

That red message is an **error**, not a disaster. Here it simply means "I don't know an object called `Age`." Read errors as hints.

::: your-turn
### Your turn {.unnumbered}

Create an object called `study_hours` and give it the value 12. Then create `study_hours_per_day` equal to `study_hours / 7` and print it.
:::

------------------------------------------------------------------------

# Operators

**Operators** are used to compare objects or numbers. They give back one or more **logical** values: either `TRUE` or `FALSE`.

These are the most common comparison operators:

| Operator | Meaning                     |
|----------|-----------------------------|
| `==`     | is equal to                 |
| `!=`     | is not equal to             |
| `>`      | is greater than             |
| `<`      | is smaller than             |
| `>=`     | is greater than or equal to |
| `<=`     | is smaller than or equal to |

For example:

```{r}
10 > 5    # is 10 greater than 5?
10 < 5    # is 10 smaller than 5?
10 == 10  # is 10 equal to 10?
```

::: {.callout-warning title="Watch out"}
To ask "is it equal?" you must use **two** equals signs `==`. A single `=` means something different and will not work here.
:::

::: your-turn
### Your turn {.unnumbered}

In your script, check whether:

(a) `7` is greater than `3`, and
(b) `4` is equal to `5`.

Run each line and read the `TRUE` / `FALSE` answer.
:::

------------------------------------------------------------------------

# Vectors: working with many values at once

In research we rarely have one number; we have many (one score per participant). A **vector** is an ordered collection of values. We build one with `c()`, which stands for *combine*. Put the values inside, separated by commas.

```{r}
scores <- c(28, 35, 31, 40, 22)
scores
```

`scores` now holds five values in one object.

A powerful feature: arithmetic on a vector applies to **every element at once**. No loop needed.

```{r}
scores + 10   # add 10 to each score
scores * 2    # double each score
```

Vectors do not have to be numbers. A **character** vector holds text (each piece in quotes), and a **logical** vector holds `TRUE`/`FALSE` values:

```{r}
names  <- c("Sanne", "Tim", "Noor", "Mees", "Lotte")
passed <- c(TRUE, FALSE, TRUE, TRUE, FALSE)
names
passed
```

::: {.callout-warning title="Watch out"}
Every value needs its own comma, and text needs quotes. `c("Sanne" "Tim")` (missing comma) gives an error; so does `c(Sanne, Tim)` (missing quotes -- R would look for objects named Sanne and Tim).
:::

::: your-turn
### Your turn {.unnumbered}

Create a vector `engagement` with the values 4, 5, 3, 5, 2 (imagine these are study-engagement ratings). Then print its `mean()` and its `length()`.
:::

------------------------------------------------------------------------

# Picking out values (indexing)

Often you want just one value, or a few, out of a vector. You do this with **square brackets** `[ ]` and a position number. **R counts from 1**, so the first element is position 1.

```{r}
scores[1]   # the first score
scores[3]   # the third score
```

Pick several at once by putting a vector of positions inside:

```{r}
scores[c(1, 2)]   # the first and second scores
```

A **minus** sign means "everything except":

```{r}
scores[-1]   # all scores except the first
```

Most powerful of all: put a **condition** inside the brackets to keep only the values that match. This is how you filter data.

```{r}
scores > 30          # TRUE/FALSE for each score
scores[scores > 30]  # keep only the scores above 30
```

Read the last line as: "give me the elements of `scores` where `scores` is greater than 30."

::: {.callout-tip collapse="true" title="Going further (optional)"}
Combine conditions with `&` ("and") or `|` ("or"), e.g. `scores[scores > 25 & scores < 40]`.
:::

::: your-turn
### Your turn {.unnumbered}

From `scores`, (a) print the last value using its position number, and (b) keep only the scores below 30.

*(Hint for (a): `scores` has 5 values.)*
:::

------------------------------------------------------------------------

# Functions

Functions have the form `name(input)`. They take an input, apply operations, and give back an output. A function is a ready-made command: you write its name and put the input in round brackets. Many useful **functions** take a vector and summarise it:

```{r}
length(scores)   # how many values?
sum(scores)      # add them all up
mean(scores)     # the average
min(scores)      # the smallest
max(scores)      # the largest
```

::: your-turn
### Your turn {.unnumbered}

Use the `engagement` vector you made earlier. Print its `sum()`, its `max()`, and its `min()`.
:::

------------------------------------------------------------------------

# Borrowing extra tools: packages and libraries

Base R already does a lot. **Packages** add extra ready-made tools written by others. Two steps, and people mix them up constantly:

-   **Install** a package: download it to your computer. You do this **once**.
-   **Load** a package with `library()`: switch it on. You do this **every session** you want to use it.

::: {.callout-note title="Analogy"}
Installing a package is like buying a book once and putting it on your shelf. `library()` is taking that book off the shelf each time you need it. Buying it again every day would be silly; so is re-installing.
:::

The syntax (you do not need to run this today -- we will install specific packages when the course needs them):

```{r}
#| eval: false
install.packages("psych")   # download once (note: name in quotes)
library(psych)              # switch on each session (no quotes here)
```

::: {.callout-warning title="Watch out"}
`install.packages()` needs the name **in quotes**; `library()` does **not**. And if you ever see `Error: there is no package called ...`, it means the package was never installed.
:::

------------------------------------------------------------------------

# Saving your work

Save your R script by clicking the save symbol (the small floppy disk), or by pressing *Ctrl* and *S* on Windows (*Cmd* and *S* on Mac). You can also click *File* \> *Save*.

The first time you save a new script, RStudio asks you where to put it. Choose your `research_methods` folder and give the file a clear name, for example `week1.R`.

------------------------------------------------------------------------

# Recap: what you can now do

You started today with no programming experience. You can now:

-   open RStudio and find your way around the four panes;
-   set the working directory;
-   create, write in, and save an R script;
-   use R as a calculator and write comments;
-   store values in objects and follow R's naming rules;
-   compare values with operators and read `TRUE` / `FALSE` answers;
-   build vectors with `c()` and do maths on all values at once;
-   pick out values with `[ ]`, including by a condition;
-   use functions such as `mean()` and `length()`;
-   explain the difference between installing and loading a package.

A small cheat sheet to keep:

| You want to...           | You write...               |
|--------------------------|----------------------------|
| Store a value            | `x <- 5`                   |
| Make a vector            | `v <- c(1, 2, 3)`          |
| Average of a vector      | `mean(v)`                  |
| Number of elements       | `length(v)`                |
| Compare two values       | `a == b`, `a > b`          |
| Get the 2nd element      | `v[2]`                     |
| Keep elements above 3    | `v[v > 3]`                 |
| Add a comment            | `# like this`              |
| Install a package (once) | `install.packages("name")` |
| Load a package (session) | `library(name)`            |

Well done. Keep your script and your `research_methods` folder -- we build directly on them next week.

------------------------------------------------------------------------
